---
title: "Exploring Demographic Factors in College Scorecard Data"
author: "Julia Silge"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: pygments
    theme: paper
    toc: yes
---

```{r, echo = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, dpi = 180)
options(width=100, dplyr.width = 150)
```

Let's explore and understand better the variables in the [College Scorecard data]((https://collegescorecard.ed.gov/data/)) that are related to *student demographics*.

## Finding the Demographic Variables

Let's open up the version of the College Scorecard data Jacob made and remove the rows that are all `NA`.

```{r}
library(readr)
colleges <- read_csv("data/scorecard_reduced_bachelors.csv", na = c("NULL", "PrivacySuppressed"))
dim(colleges)
```

Now let's open up the data dictionary.

```{r}
dictionary <- read_csv("data/CollegeScorecardDataDictionary-09-08-2015.csv")
dim(dictionary)
```

The data dictionary will tell us which variables are the demographic variables. Let's check out what we have in the data dictionary.

```{r}
names(dictionary)
```

It is `dev-category` that tells us which variables belong to which category. What are the categories?

```{r}
levels(factor(dictionary$`dev-category`))
```

Which variables belong to `student`?

```{r}
library(dplyr)
student_names <- dictionary %>% 
    filter(`dev-category` %in% c("student")) %>% 
    select(`VARIABLE NAME`, `NAME OF DATA ELEMENT`)

student_names
```

## Understanding the Student Data

What are these 94 aid variables?

```{r}
student_names$`NAME OF DATA ELEMENT`
```

## Variables for Diversity

Let's look at the total share of enrollment of undergraduate degree-seeking students who identify with certain racial/ethnic groups. How many schools are missing data?

```{r}
colleges %>% 
    summarize(`NA white` = mean(is.na(UGDS_WHITE)),
              `NA black` = mean(is.na(UGDS_BLACK)),
              `NA Hispanic` = mean(is.na(UGDS_HISP)),
              `NA Asian` = mean(is.na(UGDS_ASIAN)))
```

These are obviously the same schools since these proportions match exactly. What kind of schools are these?

```{r}
colleges %>% 
        filter(sch_deg == 3 & is.na(UGDS_WHITE)) %>% 
        select(INSTNM)
```

## More Student Demographics

Let's look at student characteristics like the percentage of undergraduates aged 25 and above and the percentage first-generation students.

```{r}
colleges %>% 
    summarize(`NA over 25` = mean(is.na(UG25abv)),
              `NA 1st gen` = mean(is.na(PAR_ED_PCT_1STGEN)))
```

Those variables for students over 25 and 1st generation students look useable.

## Summary

* I suggest for the student demographic data we use `UGDS_WHITE`, `UGDS_BLACK`, `UGDS_HISP`, and `UGDS_ASIAN` (the total share of enrollment of undergraduate degree-seeking students who identify with given racial/ethnic groups).
* I suggest we also use `PAR_ED_PCT_1STGEN` and `UG25abv` to include information about first-generation students and students over 25.
* Jacob's "folded" version of the College Scorecard data for four-year institutions that substitutes older years for missing data seems to be working very well.
