---
title: "Recommendation by cosine similarity"
author: "Krishanu Banerjee"
date: "Tuesday, September 27, 2016"
output: word_document
---

 Objective is to find similarity between institutions so that application can recommend which insitutions are the best fit for a particular student. Students will provide his/her background information and application will find the best fit institution for the student based on cosine similarity.
 
 
Data preperation- Attributes are based on mainly students background, family background, interest of study and cost of study.
The name of the institution and zip code combination are considered as unique key.Cost for the program is selected based on private, public or others. Multiple years of data averaged over unique key.

```{r}
library(RSQLite)
library('magrittr')  
library('tidyr')     
library('dplyr')     
library('ggplot2')   
library('gridExtra') 
library('bnlearn')   
library('leaflet')   
library('htmltools') 
library('RColorBrewer')
library("gplots")
library("cluster")


db <- dbConnect(dbDriver("SQLite"), "D:/college_score/output/database.sqlite")


trainSummary<- dbGetQuery(db, "

                   select LOWER(INSTNM) || '-' || ZIP AS INSTNM_ZIP
                          ,AVG(SAT_AVG_ALL)
                          ,CASE WHEN COSTT4_A = 'NA' THEN AVG(COSTT4_P) ELSE  AVG(COSTT4_A) END as COSTT --cost anual academic year 
                           ,AVG(PCTFLOAN) --federal loan rate
                          ,AVG(UG25abv) -- undergrad over age 25
                          ,AVG(PAR_ED_N)
                          ,AVG(PAR_ED_PCT_1STGEN)
                          ,AVG(DEBT_MDN) -- median debt
            ,AVG(gt_25k_p6) -- shareof students earning over 25k after 6 year
                          ,AVG(NONCOM_RPY_5YR_RT)
                          ,AVG(first_gen)
                          , AVG(md_faminc)
                          ,AVG(pct_ba)
                          , AVG(pct_grad_prof)
                          , AVG(median_hh_inc)
                          , AVG(unemp_rate)
                          ,AVG(loan_ever)
                           ,AVG(PCIP01)
 ,AVG(PCIP03)
 ,AVG(PCIP04)
 ,AVG(PCIP05)
 ,AVG(PCIP09)
 ,AVG(PCIP10)
 ,AVG(PCIP11)
 ,AVG(PCIP12)
 ,AVG(PCIP13)
 ,AVG(PCIP14)
 ,AVG(PCIP15)
 ,AVG(PCIP16)
 ,AVG(PCIP19)
 ,AVG(PCIP22)
 ,AVG(PCIP23)
 ,AVG(PCIP24)
 ,AVG(PCIP25)
 ,AVG(PCIP26)
 ,AVG(PCIP27)
 ,AVG(PCIP29)
 ,AVG(PCIP30)
 ,AVG(PCIP31)
 ,AVG(PCIP38)
 ,AVG(PCIP39)
 ,AVG(PCIP40)
 ,AVG(PCIP41)
 ,AVG(PCIP42)
 ,AVG(PCIP43)
 ,AVG(PCIP44)
 ,AVG(PCIP45)
 ,AVG(PCIP46)
 ,AVG(PCIP47)
 ,AVG(PCIP48)
 ,AVG(PCIP49)
 ,AVG(PCIP50)
 ,AVG(PCIP51)
 ,AVG(PCIP52)
 ,AVG(PCIP54)
,ifnull(case when (NPT4_OTHER is null and NPT4_PRIV is null and NPT4_PROG is null) then  AVG(NPT4_PUB)
      when (NPT4_OTHER is null and  NPT4_PUB is null and NPT4_PROG is null) then  AVG(NPT4_PRIV)
      when (NPT4_OTHER is null and  NPT4_PUB is null and NPT4_PRIV is null) then  AVG(NPT4_PROG)
      when (NPT4_PROG is null and  NPT4_PUB is null and NPT4_PRIV is null) then  AVG(NPT4_OTHER) end,0) NPT4
,ifnull(case when(NPT41_OTHER is null and NPT41_PRIV is null and NPT41_PROG is null) then  AVG(NPT41_PUB)
      when(NPT41_OTHER is null and NPT41_PUB is null and NPT41_PROG is null) then  AVG(NPT41_PRIV)
      when(NPT41_OTHER is null and NPT41_PUB is null and NPT41_PRIV is null) then  AVG(NPT41_PROG)
    when(NPT41_PROG is null and NPT41_PUB is null and NPT41_PRIV is null) then  AVG(NPT41_OTHER) end,0) NPT41  
,ifnull(case when(NPT42_OTHER is null and NPT42_PRIV is null and NPT42_PROG is null) then  AVG(NPT42_PUB)
      when(NPT42_OTHER is null and NPT42_PUB is null and NPT42_PROG is null) then  AVG(NPT42_PRIV)
      when(NPT42_OTHER is null and NPT42_PUB is null and NPT42_PRIV is null) then  AVG(NPT42_PROG)
    when(NPT42_PROG is null and NPT42_PUB is null and NPT42_PRIV is null) then  AVG(NPT42_OTHER) end,0) NPT42	  
,ifnull(case when(NPT43_OTHER is null and NPT43_PRIV is null and NPT43_PROG is null) then  AVG(NPT43_PUB)
      when(NPT43_OTHER is null and NPT43_PUB is null and NPT43_PROG is null) then  AVG(NPT43_PRIV)
      when(NPT43_OTHER is null and NPT43_PUB is null and NPT43_PRIV is null) then  AVG(NPT43_PROG)
	  when(NPT43_PROG is null and NPT43_PUB is null and NPT43_PRIV is null) then  AVG(NPT43_OTHER) end,0) NPT43	  
,ifnull(case when(NPT44_OTHER is null and NPT44_PRIV is null and NPT44_PROG is null) then  AVG(NPT44_PUB)
      when(NPT44_OTHER is null and NPT44_PUB is null and NPT44_PROG is null) then  AVG(NPT44_PRIV)
      when(NPT44_OTHER is null and NPT44_PUB is null and NPT44_PRIV is null) then  AVG(NPT44_PROG)
	  when(NPT44_PROG is null and NPT44_PUB is null and NPT44_PRIV is null) then  AVG(NPT44_OTHER) end,0) NPT44	 
,ifnull(case when(NPT45_OTHER is null and NPT45_PRIV is null and NPT45_PROG is null) then  AVG(NPT45_PUB)
      when(NPT45_OTHER is null and NPT45_PUB is null and NPT45_PROG is null) then  AVG(NPT45_PRIV)
      when(NPT45_OTHER is null and NPT45_PUB is null and NPT45_PRIV is null) then  AVG(NPT45_PROG)
	  when(NPT45_PROG is null and NPT45_PUB is null and NPT45_PRIV is null) then  AVG(NPT45_OTHER) end,0) NPT45		  
,ifnull(case when(NPT4_048_OTHER is null and NPT4_048_PRIV is null and NPT4_048_PROG is null) then  AVG(NPT4_048_PUB)
      when(NPT4_048_OTHER is null and NPT4_048_PUB is null and NPT4_048_PROG is null) then  AVG(NPT4_048_PRIV)
      when(NPT4_048_OTHER is null and NPT4_048_PUB is null and NPT4_048_PRIV is null) then  AVG(NPT4_048_PROG)
	  when(NPT4_048_PROG is null and NPT4_048_PUB is null and NPT4_048_PRIV is null) then  AVG(NPT4_048_OTHER) end,0) NPT4_048	    
,ifnull(case when(NPT4_3075_OTHER is null and NPT4_3075_PRIV is null and NPT4_3075_PROG is null) then  AVG(NPT4_3075_PUB)
      when(NPT4_3075_OTHER is null and NPT4_3075_PUB is null and NPT4_3075_PROG is null) then  AVG(NPT4_3075_PRIV)
      when(NPT4_3075_OTHER is null and NPT4_3075_PUB is null and NPT4_3075_PRIV is null) then  AVG(NPT4_3075_PROG)
	  when(NPT4_3075_PROG is null and NPT4_3075_PUB is null and NPT4_3075_PRIV is null) then  AVG(NPT4_3075_OTHER) end,0) NPT4_3075	      	  
,ifnull(case when(NPT4_75UP_OTHER is null and NPT4_75UP_PRIV is null and NPT4_75UP_PROG is null) then  AVG(NPT4_75UP_PUB)
      when(NPT4_75UP_OTHER is null and NPT4_75UP_PUB is null and NPT4_75UP_PROG is null) then  AVG(NPT4_75UP_PRIV)
      when(NPT4_75UP_OTHER is null and NPT4_75UP_PUB is null and NPT4_75UP_PRIV is null) then  AVG(NPT4_75UP_PROG)
	  when(NPT4_75UP_PROG is null and NPT4_75UP_PUB is null and NPT4_75UP_PRIV is null) then  AVG(NPT4_75UP_OTHER) end,0) NPT4_75UP

                       FROM Scorecard 
                          
                          WHERE
                          SCH_DEG=3 and
                          
                          LOWER(INSTNM) IN 
                          
                          (select LOWER(INSTNM) 
                          
                          from Scorecard 
                          
                          group by LOWER(INSTNM),ZIP
                          
                          HAVING (SUM(CASE WHEN SAT_AVG_ALL IS NULL THEN 0 ELSE 1 END) > 0))
                          
                          GROUP BY LOWER(INSTNM) || '-' || ZIP")
```
null values replaced by mean of the field.

```{r}
for(i in 1:ncol(trainSummary)){
  trainSummary[,i][is.na(trainSummary[,i])] <- mean(trainSummary[,i], na.rm = TRUE)
}

```
Cosine similarity function 

```{r}
getCosine<-function(x,y)
  {
  this.cosine=sum(x*y)/(sqrt(sum(x*x))*sqrt(sum(y*y)))
  return (this.cosine)
  }

```
Initial testing of the function

```{r}
institution= trainSummary[,1]
trainSum=trainSummary[-c(1)]
insti_916=trainSum[916,]
insti_2=trainSum[2,]
insti_18=trainSum[18,]
test_inst=trainSum[4977,]
getCosine(insti_2,insti_2)
getCosine(insti_2,insti_18)

```
Testing for a average student of 'Yale university'(tested both for the best and the worst match)

```{r}
sim_vec=c()
for (i in 1:nrow(trainSum)){
     sim_vec[i]=getCosine(test_inst,trainSum[i,])
}
res_data=NULL
res_data=data.frame(cbind(institution,sim_vec))

head(res_data[order(-sim_vec),],20)
head(res_data[order(sim_vec),],10)
```

Conclusion-

1) For this particular testing we can found some similar university as 'Yale' but not all.
2) Other relevant fields need to consider.
3) Null value replacement with mean is very simplistic assumption. Need better approach.
4) Need to try other similarity methods, such as Pearson correlation, Euclidean,Bayesian etc.

```{r, echo=FALSE}

```


